---
output: github_document
---

<!-- README.md is generated from README.Rmd. -->

# RClinVarbitration

`RClinVarbitration` streams official ClinVar VCV XML/XML.GZ into focused
relational DuckDB tables. It retains public VCV, RCV, SCV, AlleleID, and
VariationID identifiers instead of exposing XML parser coordinates.

The package-owned native extension uses a libxml2 forward reader. An import
reads the XML exactly once into a compact temporary semantic-fact table,
projects ordinary ClinVar relations, and drops the temporary table. It does not
build an XML DOM or R data frame and does not persist generic XML nodes, edges,
or element statements.

Because the extension uses DuckDB's `C_STRUCT_UNSTABLE` interface, the package
bundles and selects exact-version artifacts for DuckDB `v1.5.0` through
`v1.5.4`. Connections must allow locally built unsigned extensions.

## Relational schema

| Relation | One row per |
|:--|:--|
| `clinvar_variants` | VCV variation archive |
| `clinvar_alleles` | allele or variation component |
| `clinvar_locations` | assembly-specific allele location |
| `clinvar_genes` | allele–gene relation |
| `clinvar_rcv_assertions` | RCV aggregate assertion |
| `clinvar_scv_assertions` | submitted SCV assertion |
| `clinvar_conditions` | condition attached to a VCV, RCV, or SCV |
| `clinvar_condition_names` | preferred or alternate condition name |
| `clinvar_observations` | SCV observation |
| `clinvar_citations` | attributable citation |
| `clinvar_citation_identifiers` | PMID, DOI, or other citation identifier |
| `clinvar_attributes` | typed source attribute retained with its context |
| `clinvar_text` | attributable comment, definition, condition name, or other discovery text |
| `clinvar_normalized_alleles` | assembly/locus/ref/alt join view |

The source package includes one **unaltered official NCBI VCV XML.GZ record**
for a fast executable smoke test. It is not synthetic data and is not presented
as a full-release benchmark. Full releases use the same import call against
`ClinVarVCVRelease_00-latest.xml.gz`; they belong in a file-backed DuckDB and
are intentionally not reparsed whenever this README is rendered.

```{r example}
library(DBI)
library(duckdb)
library(RClinVarbitration)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rclinvarbitration_enable(con)
clinvar_xml <- system.file(
  "extdata", "VCV_XML_VCV000091629.xml.gz",
  package = "RClinVarbitration", mustWork = TRUE
)
counts <- rclinvarbitration_import_xml(
  con,
  clinvar_xml,
  release_id = "ncbi-vcv-91629"
)
knitr::kable(
  data.frame(relation = names(counts), rows = unname(counts)),
  row.names = FALSE
)
```

The normalized allele view is the join surface for downstream
Rduckhts/DuckHTS work. RClinVarbitration does not duplicate VEP or PM5.

```{r alleles}
alleles <- dbGetQuery(con, "
  SELECT vcv_accession, allele_id, variation_id, assembly,
         chromosome, position_vcf, reference, alternate, canonical_spdi
  FROM clinvar_normalized_alleles
  ORDER BY assembly
")
knitr::kable(alleles, row.names = FALSE)
```

SCV classifications remain separate from ClinVar's VCV and RCV aggregates, so
later policy SQL can filter submitters and dates without reconstructing
assertions from XML paths.

```{r assertions}
assertions <- dbGetQuery(con, "
  SELECT scv_accession, submitter_name, classification,
         review_status, date_last_evaluated
  FROM clinvar_scv_assertions
  ORDER BY scv_accession
")
knitr::kable(assertions, row.names = FALSE)
```

Long assertion explanations are queried as attributable text. The README
selects and truncates them in SQL so knitr never prints a data frame padded to
the width of the longest ClinVar comment.

```{r evidence}
evidence <- dbGetQuery(con, "
  SELECT s.scv_accession,
         left(t.text, 180) || CASE WHEN length(t.text) > 180 THEN '…' ELSE '' END AS excerpt
  FROM clinvar_text t
  JOIN clinvar_scv_assertions s
    ON s.release_id = t.release_id
   AND s.assertion_entity_id = t.scv_entity_id
  WHERE t.section = 'comment'
  ORDER BY s.scv_accession
")
knitr::kable(evidence, row.names = FALSE)

DBI::dbDisconnect(con, shutdown = TRUE)
```

For a full release, open a file-backed connection and pass the local official
XML.GZ path to `rclinvarbitration_import_xml()`. The import is one forward XML
pass; the resulting DuckDB relations are queried repeatedly without reparsing
the gzip file.

ClinVarbitration decision rules are intentionally a later versioned SQL layer
over `clinvar_scv_assertions`. P/LP policy outputs will join
`clinvar_normalized_alleles` by assembly, chromosome, position, reference, and
alternate allele.

## Acknowledgement

[Teague Sterling's DuckDB Webbed extension](https://github.com/teaguesterling/duckdb_webbed)
informed the initial assessment of SQL-native XML ingestion. RClinVarbitration
does not vendor or execute Webbed: its package-owned libxml2 scan handles the
official gzip VCV release and is built against each supported exact DuckDB
engine version.
