---
output: github_document
---

<!-- README.md is generated from README.Rmd. -->

# RClinVarbitration

`RClinVarbitration` streams the complete official ClinVar VCV XML/XML.GZ
release into focused relational DuckDB tables. It retains public VCV, RCV, SCV,
AlleleID, VariationID, and disease identifiers instead of exposing XML parser
coordinates.

The package-owned native extension uses a libxml2 forward reader. An import
reads the gzip XML exactly once and writes one compact JSON-backed row per
selected ClinVar entity to a disk-backed staging table. It then projects
ordinary ClinVar relations without an EAV pivot and drops staging. It does not
build an XML DOM or R data frame and does not persist generic XML nodes, edges,
or element statements.

Because the extension uses DuckDB's `C_STRUCT_UNSTABLE` interface, the package
bundles and selects exact-version artifacts for DuckDB `v1.5.0` through
`v1.5.4`. Connections must allow locally built unsigned extensions.

## Relational schema

| Relation | One row per |
|:--|:--|
| `clinvar_variants` | VCV variation archive |
| `clinvar_alleles` | allele or variation component |
| `clinvar_locations` | assembly-specific allele location |
| `clinvar_genes` | allele–gene relation |
| `clinvar_rcv_assertions` | disease-specific RCV aggregate assertion |
| `clinvar_scv_assertions` | submitted SCV assertion |
| `clinvar_conditions` | condition attached to an RCV or SCV |
| `clinvar_condition_names` | preferred or alternate condition name |
| `clinvar_observations` | SCV observation |
| `clinvar_citations` | SCV-attributable citation |
| `clinvar_citation_identifiers` | PMID, DOI, or other citation identifier |
| `clinvar_attributes` | typed SCV source attribute retained with its context |
| `clinvar_text` | attributable assertion comment, condition name, or evidence text |
| `clinvar_normalized_alleles` | assembly/locus/ref/alt join view |
| `clinvar_disease_aggregates` | VCV/allele × RCV × disease aggregate |
| `clinvar_disease_submissions` | VCV/allele × SCV × disease submission |

The release-scale tables retain explicit logical key columns but do not create
DuckDB ART indexes: those indexes must remain memory-resident and are a poor fit
for a complete analytical release. The small `clinvar_releases` completion
catalogue retains its primary key.

The executable README requires the complete local
`ClinVarVCVRelease_00-latest.xml.gz`, selected with
`CLINVAR_VCV_XML_FILE`, and an explicit immutable release label in
`CLINVAR_RELEASE_ID`. It does not substitute the bundled one-record regression
fixture. A **file-backed** DuckDB holds the complete import so every following
query reuses the same data rather than rescanning XML. Set
`CLINVAR_DUCKDB_FILE` to retain and reuse a completed import across renders;
otherwise the executable document uses a temporary database.

```{r import-full-release}
library(DBI)
library(duckdb)
library(RClinVarbitration)

clinvar_xml <- Sys.getenv("CLINVAR_VCV_XML_FILE")
release_id <- Sys.getenv("CLINVAR_RELEASE_ID")
stopifnot(
  nzchar(release_id),
  file.exists(clinvar_xml),
  identical(basename(clinvar_xml), "ClinVarVCVRelease_00-latest.xml.gz")
)

clinvar_db <- Sys.getenv("CLINVAR_DUCKDB_FILE")
persistent_db <- nzchar(clinvar_db)
if (persistent_db) {
  dir.create(dirname(clinvar_db), recursive = TRUE, showWarnings = FALSE)
  clinvar_db <- normalizePath(clinvar_db, mustWork = FALSE)
} else {
  clinvar_db <- tempfile("rclinvarbitration-readme-", fileext = ".duckdb")
}
clinvar_tmp <- paste0(clinvar_db, ".tmp")
con <- dbConnect(duckdb(
  dbdir = clinvar_db,
  config = list(
    allow_unsigned_extensions = "true",
    memory_limit = Sys.getenv("CLINVAR_DUCKDB_MEMORY_LIMIT", "2GB"),
    preserve_insertion_order = "false",
    temp_directory = clinvar_tmp,
    threads = "2"
  )
))
rclinvarbitration_enable(con)
rclinvarbitration_init(con)
release_sql <- as.character(dbQuoteString(con, release_id))
stored_release <- dbGetQuery(con, paste0(
  "SELECT source_path FROM clinvar_releases WHERE release_id = ", release_sql
))
count_relations <- c(
  variants = "clinvar_variants",
  alleles = "clinvar_alleles",
  rcv_assertions = "clinvar_rcv_assertions",
  scv_assertions = "clinvar_scv_assertions",
  conditions = "clinvar_conditions",
  observations = "clinvar_observations",
  citations = "clinvar_citations",
  text = "clinvar_text"
)
if (nrow(stored_release)) {
  stopifnot(identical(stored_release$source_path, normalizePath(clinvar_xml)))
  counts <- vapply(count_relations, function(table) {
    dbGetQuery(con, paste0(
      "SELECT count(*) AS n FROM ", table, " WHERE release_id = ", release_sql
    ))$n[[1L]]
  }, numeric(1))
} else {
  counts <- rclinvarbitration_import_xml(
    con,
    clinvar_xml,
    release_id = release_id
  )
}
knitr::kable(
  data.frame(relation = names(counts), rows = unname(counts)),
  row.names = FALSE
)
```

## Disease-specific aggregates

RCVs are the disease-specific ClinVar aggregates. They remain distinct from the
VCV-level aggregate and are joined directly to their classified condition. This
is the primary substrate for reproducing ClinVarbitration and for deriving
alternative disease-aware policies.

```{r disease-aggregate-summary}
disease_summary <- dbGetQuery(con, "
  SELECT aggregate_classification, aggregate_review_status,
         count(*) AS disease_aggregates,
         count(DISTINCT disease_identifier) AS identified_diseases
  FROM clinvar_disease_aggregates
  GROUP BY aggregate_classification, aggregate_review_status
  ORDER BY disease_aggregates DESC
  LIMIT 20
")
knitr::kable(disease_summary, row.names = FALSE)
```

```{r disease-aggregate-examples}
disease_aggregates <- dbGetQuery(con, "
  SELECT vcv_accession, rcv_accession,
         disease_database, disease_identifier, disease_name,
         aggregate_classification, aggregate_review_status
  FROM clinvar_disease_aggregates
  WHERE disease_identifier IS NOT NULL
  ORDER BY vcv_accession, rcv_accession
  LIMIT 12
")
knitr::kable(disease_aggregates, row.names = FALSE)
```

Individual SCVs retain their disease mappings, submitters, dates, review status,
and contribution flags. Policy SQL can therefore recompute an aggregate per
allele and disease instead of accepting only ClinVar's top-line VCV label.

```{r disease-submission-summary}
submission_summary <- dbGetQuery(con, "
  SELECT classification, review_status,
         count(*) AS disease_submissions,
         count(DISTINCT submitter_id) AS submitters
  FROM clinvar_disease_submissions
  GROUP BY classification, review_status
  ORDER BY disease_submissions DESC
  LIMIT 20
")
knitr::kable(submission_summary, row.names = FALSE)
```

## Normalized allele join surface

The normalized allele view is the join surface for downstream
Rduckhts/DuckHTS work. RClinVarbitration does not duplicate VEP or PM5.

```{r allele-summary}
alleles <- dbGetQuery(con, "
  SELECT assembly,
         count(*) AS normalized_alleles,
         count(DISTINCT allele_id) AS allele_ids
  FROM clinvar_normalized_alleles
  GROUP BY assembly
  ORDER BY assembly
")
knitr::kable(alleles, row.names = FALSE)
```

Long assertion explanations remain attributable to the source SCV. The README
truncates selected text in SQL so knitr does not pad console output to the width
of the longest ClinVar comment.

```{r evidence}
evidence <- dbGetQuery(con, "
  SELECT s.scv_accession, s.submitter_name,
         left(t.text, 180) || CASE WHEN length(t.text) > 180 THEN '…' ELSE '' END AS excerpt
  FROM clinvar_text t
  JOIN clinvar_scv_assertions s
    ON s.release_id = t.release_id
   AND s.assertion_entity_id = t.scv_entity_id
  WHERE t.section = 'comment'
  ORDER BY s.scv_accession
  LIMIT 8
")
knitr::kable(evidence, row.names = FALSE)

DBI::dbDisconnect(con, shutdown = TRUE)
if (!persistent_db) {
  unlink(
    c(clinvar_db, paste0(clinvar_db, ".wal"), clinvar_tmp),
    recursive = TRUE,
    force = TRUE
  )
}
```

ClinVarbitration decision rules are a versioned SQL layer over
`clinvar_disease_submissions`. P/LP outputs join
`clinvar_normalized_alleles` by assembly, chromosome, position, reference, and
alternate allele. DuckLake can later retain each merged release as a snapshot,
but XML ingestion and policy semantics remain explicit here.

## Acknowledgement

[Teague Sterling's DuckDB Webbed extension](https://github.com/teaguesterling/duckdb_webbed)
informed the initial assessment of SQL-native XML ingestion. RClinVarbitration
does not vendor or execute Webbed: its package-owned libxml2 scan handles the
official gzip VCV release and is built against each supported exact DuckDB
engine version.
